
---
title: "Team_TAM_Case_Study02"
authors: 
- Andrew Heroy <aheroy@smu.edu>
- Martin Garcia <marting@smu.edu>
- Tej Tenmattam <ttenmattam@smu.edu>
date: "11/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# MSDS 6306: Doing Data Science
## Case Study 02
#### 1. Load and clean the Data: 
```{r}
# Install and load the library readxl so you can load the dataset
# install.packages("readxl")
library(readxl)
# read the dataset
attrition_data <-read_excel("../data/CaseStudy2-data.xlsx")
# describe the dataset
str(attrition_data)
# Clean the data
# Over18 variable is "Y" for all the employees, there is no variability so we will remove it.
# EmployeeCount is "1" for all the employees, there is no variability so we will remove it.
# StandardHours is "80" for all the employees, there is no variability so we will remove it.
# EmployeeNumber is an assigned number for each employee, this does not cause any variability so we will remove it.
attrition_data$Over18 <- NULL
attrition_data$EmployeeCount <- NULL
attrition_data$StandardHours <- NULL
attrition_data$EmployeeNumber<-NULL

#Calculating the number of missing values for each variable
NA_sum <- sort(sapply(attrition_data, function(x) sum(is.na(x))), decreasing = TRUE)
print(NA_sum) 
# There are no missing values for each of the variables.
```
#### 2. EDA through Correlation Matrix:
```{r}
# Identify the numeric variables in the dataset
numeric <- sapply(attrition_data, is.numeric)
print(numeric)

# Separate the numeric variables into a separate dataset.
numeric_dat <- attrition_data[,numeric==TRUE]

#install.packages("fastDummies")
library(fastDummies)

# Generate a heatmap on the numeric data
numeric_attr <- cbind(numeric_dat[1:1470,], attrition_data['Attrition'])
numeric_attr_dummy <- fastDummies::dummy_cols(numeric_attr, select_columns = "Attrition")
numeric_attr_dummy$Attrition <-NULL
numeric_attr_dummy$Attrition_No <-NULL
corr_df <- cor(numeric_attr_dummy)
#install.packages("corrplot")
library(corrplot)
library(ggplot2)
highcor <- as.matrix(sort(corr_df[ ,'Attrition_Yes'], decreasing = TRUE))
corr.idx <- names(which(apply(highcor, 1, function(x) (x > 0.05 | x < -0.05))))
corrplot(as.matrix(corr_df[corr.idx,corr.idx]), method='square', 
         addCoef.col = 'black', tl.cex = .8,cl.cex = .8, number.cex=.6)
# Based on the correlation matrix: Total Working Years,Job Level,Years in Current Role,Monthly Income and Age are the variables that are highly correlated with attrition.
```
#### 3. EDA through Scatterplot Matrix:
```{r}
# Scatter Plot Matrix
pairs(~ Age+DailyRate+DistanceFromHome+HourlyRate+MonthlyIncome+MonthlyRate+NumCompaniesWorked+PercentSalaryHike+TotalWorkingYears+YearsAtCompany+YearsInCurrentRole+YearsSinceLastPromotion,data=numeric_attr, main="Scatterplot of Important variables", cex=0.01)
# 23 different variables makes it a bit crowded and hard to visualize, so we will proceed with other EDA tools.
```

#### 4. EDA through Decision Tree Model:
```{r}
# Some of the variables are of integer data type, it makes more sense to convert them to factor data type.
to_factor <- c("Education", "EnvironmentSatisfaction", "JobInvolvement", "JobLevel", "JobSatisfaction", "PerformanceRating", "RelationshipSatisfaction", "StockOptionLevel","TrainingTimesLastYear","WorkLifeBalance")
attrition_data[, to_factor] <- lapply((attrition_data[, to_factor]), as.factor)

# Looking at the dataset there are many categorical variables, so we decided to use the R packages decision tree model (rpart and rpart.plot) for EDA. Reference: https://rpubs.com/minma/cart_with_rpart
# install.packages("rpart")
# install.packages("rpart.plot")
library(rpart)
library(rpart.plot)
attrition_decision_tree <- rpart( Attrition~.,data=attrition_data,control=rpart.control(minsplit = 10))
# Display the variable importance based on the decision tree model
attrition_decision_tree$variable.importance
# Display a plot diagram of the attrition decision tree
rpart.plot(attrition_decision_tree)
# Based on the decision tree output: Overtime, MonthlyIncome, TotalWorkingYears, HourlyRate, JobRole are the most important factors influencing the attrition rates.
```
#### 5. EDA through Stepwise Regression Model:
```{r}
# stepwise forward regression
attrition_data_sw <- fastDummies::dummy_cols(attrition_data, select_columns = "Attrition")
attrition_data_sw$Attrition <-NULL
attrition_data_sw$Attrition_No <-NULL
# install olsrr from CRAN
#install.packages("olsrr")
library('olsrr')
model <- lm(Attrition_Yes ~ ., data = attrition_data_sw)
# forward selection
#ols_step_forward_p(model)
# backward selection
#ols_step_backward_p(model)
# both
ols_step_both_p(model)
# Stepwise AIC Regression
#ols_step_both_aic(model)
# After doing our analysis through these multiple methods, we could see a pattern emerge. The top most influential variables are Over Time, Total Working Years, Monthly Income, Job Role and Age. We will explore these variables more through visualization.
```
#### 6. Visualize the data using Boxplots and Histograms:
```{r}
#install packages("dplyr")
#install packages("scales")
#install packages("reshape2")
#install packages("tidyr")
#install packages("gridExtra")

library(ggplot2)
library(dplyr)
library(scales)
library(reshape2)
library(tidyr)
library(gridExtra)

```
```{r}
#Barplot for attrition counts
ggplot(data.frame(attrition_data), aes(x=Attrition)) +
  geom_bar()
```
```{r}
#We use boxplots to see any differences between the distribution of those with attrition and without. Conditions that we must be aware of that our distributions may be skewed as there were more examples with no attrition than with.

##Boxplot for all factors that are numerical

attrition_data %>%
  select(Attrition, Age, MonthlyIncome, TotalWorkingYears,        StockOptionLevel:YearsWithCurrManager,MonthlyIncome:NumCompaniesWorked,PercentSalaryHike:RelationshipSatisfaction,DailyRate, DistanceFromHome:Education, 
  EnvironmentSatisfaction, HourlyRate:JobLevel, JobSatisfaction) %>%
  gather(Measure, Value, -Attrition) %>%
  ggplot(aes(x = factor(Attrition), y = Value, fill=Attrition)) +
   stat_boxplot(geom ='errorbar') +
  geom_boxplot() +
  facet_wrap(~Measure, scales = "free_y")
```
```{r}
##For our 5 main factors / Overtime, MonthlyIncome, TotalWorkingYears, HourlyRate, JobRole and Age
#Notable points of interest are our MonthlyIncome and TotalWorkingYears
#MonthlyIncome - our average income of those with attrition have a smaller average income
#             - the range of income for for our groups are substantially different, most with no attrition have income     higher than our median with attrition
#TotalWorkingYears - The ranges are are different with most with attrition peaking at 20 years working
#Safe to assume that these factors run parrallel to age as the totalworking years increases our age does too, we can also expect our wages to increase

attrition_data %>%
  select(Attrition, Age, MonthlyIncome, TotalWorkingYears, HourlyRate) %>%
  gather(Measure, Value, -Attrition) %>%
  ggplot(aes(x = factor(Attrition), y = Value, fill=Attrition)) +
   stat_boxplot(geom ='errorbar') +
  geom_boxplot() +
  facet_wrap(~Measure, scales = "free_y")
```
```{r}

#Histogram per category, Attrition. This is measured for age against all numerical variables.


require(gridExtra)
a <- ggplot(attrition_data, aes(x=Age, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
b <- ggplot(attrition_data, aes(x=DailyRate, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
c <- ggplot(attrition_data, aes(x=DistanceFromHome, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
d <- ggplot(attrition_data, aes(x=Education, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
e <- ggplot(attrition_data, aes(x=EnvironmentSatisfaction, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
f <- ggplot(attrition_data, aes(x=HourlyRate, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
g <- ggplot(attrition_data, aes(x=JobInvolvement, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
h <- ggplot(attrition_data, aes(x=JobLevel, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
i <- ggplot(attrition_data, aes(x=JobSatisfaction, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
j <- ggplot(attrition_data, aes(x=MonthlyIncome, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
k <- ggplot(attrition_data, aes(x=MonthlyRate, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
l <- ggplot(attrition_data, aes(x=NumCompaniesWorked, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
m <- ggplot(attrition_data, aes(x=PercentSalaryHike, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
n <- ggplot(attrition_data, aes(x=PerformanceRating, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
o <- ggplot(attrition_data, aes(x=Education, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
p <- ggplot(attrition_data, aes(x=RelationshipSatisfaction, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
q <- ggplot(attrition_data, aes(x=StockOptionLevel, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
r <- ggplot(attrition_data, aes(x=TotalWorkingYears, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
s <- ggplot(attrition_data, aes(x=TrainingTimesLastYear, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
t <- ggplot(attrition_data, aes(x=WorkLifeBalance, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
u <- ggplot(attrition_data, aes(x=YearsAtCompany, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
v <- ggplot(attrition_data, aes(x=YearsInCurrentRole, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
w <- ggplot(attrition_data, aes(x=YearsSinceLastPromotion, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")
y <- ggplot(attrition_data, aes(x=YearsWithCurrManager, fill=Attrition)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")   
    
    
grid.arrange(a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s,t,u,v,w,y, ncol=4)   


```


```{r}
#Removing those with attrition value of yes, will dig deeper into those with attrition

attrition_dataNo <- attrition_data[attrition_data$Attrition != "Yes",]
ggplot(data = attrition_dataNo, aes(x = Age)) + geom_histogram(fill = 'skyblue2') 
ggplot(data = attrition_dataNo, aes(x = Education)) +geom_bar(fill='palegreen3')+ geom_histogram(fill = 'palegreen3')
ggplot(data = attrition_dataNo, aes(x = Gender)) + geom_histogram(fill = 'darkorange1')
ggplot(data = attrition_dataNo, aes(x = MaritalStatus)) + geom_histogram(fill = 'mediumpurple2')

ggplot(data.frame(attrition_dataNo), aes(x=Gender, fill=Gender), ) +
  geom_bar(stat="count")
ggplot(data.frame(attrition_dataNo), aes(x=MaritalStatus,  fill=MaritalStatus)) +
  geom_bar(stat="count",)
  
```

#### 7. Visualize the data for trends:
```{r}
# Display the attrition rate at the Company
library(ggplot2)
table(attrition_data$Attrition)
# There are a total of 237 employees who have left the company
ggplot(attrition_data, aes(x=Attrition, fill=Attrition)) + geom_bar()
# The impact of Overtime 
table(attrition_data$OverTime, attrition_data$Attrition)
ggplot(attrition_data, aes(OverTime, ..count.., fill = factor(Attrition))) + geom_bar(position="dodge")
# The impact of Monthly Income
# https://rpubs.com/pierrelafortune/cutdocumentation
summary(attrition_data$MonthlyIncome)
attrition_monthlyincome <- cut(attrition_data$MonthlyIncome, 10, include.lowest = TRUE, labels=c(1,2,3,4,5,6,7,8,9,10))
ggplot(attrition_data, aes(attrition_monthlyincome, ..count.., fill = factor(Attrition))) + geom_bar(position="dodge")
# As salary increases the attrition is decreasing. Lower salary is contributing to attrition
# Attrition based on the number of working years
summary(attrition_data$TotalWorkingYears)
attrition_totalworkingyears <- cut(attrition_data$TotalWorkingYears, 10, include.lowest = TRUE)
ggplot(attrition_data, aes(attrition_totalworkingyears, ..count.., fill = factor(Attrition))) + geom_bar(position="dodge")
# Attrition based on the Hourly rate
summary(attrition_data$HourlyRate)
attrition_hourlyrate<- cut(attrition_data$HourlyRate, 7, include.lowest = TRUE)
ggplot(attrition_data, aes(attrition_hourlyrate, ..count.., fill = factor(Attrition))) + geom_bar(position="dodge")
# Attrition based on the Job Role
table(attrition_data$JobRole, attrition_data$Attrition)
ggplot(attrition_data, aes(JobRole, ..count.., fill = factor(Attrition))) + geom_bar(position="dodge") + theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1))
# Attrition based on the Age
summary(attrition_data$Age)
attrition_age <- cut(attrition_data$Age, 8, include.lowest = TRUE)
ggplot(attrition_data, aes(attrition_age, ..count.., fill = factor(Attrition))) + geom_bar(position="dodge") + theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1))
```

#### 8. Summary:
To determine the factors that influence employee retainment we analyized 35 factors using exploratory data analysis.
This analysis included a decision tree model, correlation and scatterplot matices. Visual guides are provided through box plots and histograms. Through these visuals we were able to capture patterns and discovered the main factors affecting employee attrition include overtime, monthly income, total working years, job role and age. We recommend these 5 factors be improved through new processes to help with attrition.

